---
title: "NSA_Problem_2"
author: "James Buenfil"
date: "2024-02-07"
output: html_document
---

Specify dataset $S_N$ (a list of lists of integers, where each sublist is one of the observed rankings). In the case of interest, $S_N$ consists of only two rankings. Then we rename them using values $1, \ldots N$ (this makes the computation faster).
We can find the original names using vector unordermap and function original_names_SN.
```{r}
#example data, each list contains the order of the items ranked, e.g. list(2,4,1) means item 2 ranked first, item 4 ranked second, item 1 ranked third
S_N_true_names = list(list(1,2,8),list(2,7,1),list(100,2,3))

output = rename_SN(S_N_true_names)
#the renamed S_N
S_N = output$S_Nnew
#vector that can be used to take new item names to old item names using function "original_names_SN"
unordermap = output$unordermap

#lines below shows usage of original_names_SN. You can verify that S_N_original_names equals S_N_true_names
S_N_original_names = original_names_SN(S_N,unordermap)
print(S_N_original_names)

```


Do maximum likelihood estimation based on the data to find MLE $\sigma_0$ and MLE $\theta$.
```{r}
#Maximum likelihood estimation
#theta_init and max_iterations can be specified, otherwise they are set to
#default values
output = ML_estimation(S_N,theta_init = NULL,max_iterations = 20)
sigma0ML = output$sigma1
thetaML = output$theta #theta is related to p in the writeup via theta_j = ln(1/p_j)


#we can use original_names_SN to look at sigma0 using the true names
sigma0ML_original_names = as.list(unlist(original_names_SN(sigma0ML,unordermap)))
print(sigma0ML_original_names)
```


We can sample from the generalized mallows model using a specified $\theta$ (such as the one estimated from maximum likelihood estimation). We just have to specify the number of and lengths of the permutations we want.
```{r}
#We can sample from the model (specified by theta) using sample_GMM
#t specifies the lengths of the sampled rankings we want
t = c(1,2,3,3)
#specifying t in this way is saying that, we want to sample 4 rankings from the 
#model, with lengths 1,2, 3, and 3.
#max(t) must be smaller than or equal to length(theta). 
#Here, the modal permutation sigma0 is (1, 2, 3... )
#sample_GMM returns a list of lists representing the rankings (the same format as S_N).
S_sampled = sample_GMM(t = t,theta = thetaML)
print(S_sampled)

#Note that by the nature of sampling from a model with an infinite number of
#items, we may see items appear in the sampled rankings that have no
#corresponding item in S_N, the dataset: If we, as described above, are sampling
#from the model using the thetaML estimated from maximum likelihood using S_N
#(the renamed S_N), then the new items are simply the ones with names larger
#than max(S_N).
paste("New items in S_sampled are those with names larger than",max(unlist(S_N)))
```

We can compute the $\Delta_j$ between two rankings $\pi_1$ and $\pi_2$. Below, the rankings are random permutation of 1,2, ... 2000, so there are a large number of total inversions between the two rankings.
```{r}
n=2000
pi1 = sample(1:n, replace=FALSE)
pi2 = sample(1:n,replace=FALSE)

#pi1 = list(1,3,4)
#pi2 = list(2,4,1,7)

#we can compute the number of inversions at each rank between two rankings
deltas = compute_deltas(pi1,pi2)
print(deltas)
print(sum(deltas))
```

