---
title: "NSA_Problem_2"
author: "James Buenfil"
date: "2024-02-07"
output: html_document
---


Specify dataset $S_N$, then we rename them using values $1, \ldots N$ (this makes the computation faster).
Can find the original names using vector unordermap and function original_names_SN.
```{r}
#example data, each list contains the order of the items ranked, e.g. list(2,4,1) means item 2 ranked first, item 4 ranked second, item 1 ranked third
S_N_true_names = list(list(1,2,8),list(2,7,1),list(100,2,3))

output = rename_SN(S_N_true_names)
#the renamed S_N
S_N = output$S_Nnew
#vector that can be used to take new item names to old item names using function "original_names_SN"
unordermap = output$unordermap

#lines below shows usage of original_names_SN. You can verify that S_N_orginal_names equals S_N_true_names
#S_N_orginal_names = original_names_SN(S_N,unordermap)
#print(S_N_orginal_names)

```


Do maximum likelihood estimation based on the data to find MLE $\sigma_0$ and MLE $\theta$.
```{r}
#Maximum likelihood estimation
#theta_init and max_iterations can be specified, otherwise they are set to
#default values
output = ML_estimation(S_N,theta_init = NULL,max_iterations = 20)
sigma0ML = output$sigma1
thetaML = output$theta

#we can use original_names_SN to look at sigma0 using the true names
sigma0ML_original_names = as.list(unlist(original_names_SN(sigma0ML,unordermap)))
print(sigma0ML_original_names)
```

We can sample from the generalized mallows model using a specified $\theta$ (such as the one estimated from maximum likelihood estimation). We just have to specify the number of and lengths of the permutations we want.
```{r}
#We can sample from the model (specified by theta) using sample_GMM
#t specifies the lengths of the sampled rankings we want
t = c(1,2,3,3)
#max(t) must be smaller than or equal to length(theta). 
#Here, sigma0 is 1, 2, ...
#Note that by the nature of sampling from a model with an infinite number of
#items, we may see items appear in the sampled rankings that have no
#corresponding item in S_N, the dataset.
#However, we can still compute statistics using samples from the
#estimated maximum likelihood parameters.
#sample_GMM returns a list of rankings, the same format as S_N
S_sampled = sample_GMM(t = t,theta = thetaML)
```

We can compute the $\Delta_j$ between two rankings $\pi_1$ and $\pi_2$.
```{r}
pi1 = list(1,3,4)
pi2 = list(2,4,1,7)

#we can compute the number of inversions at each rank between two rankings
deltas = compute_deltas(pi1,pi2)
print(deltas)
```

